Learning to communicate using recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex
merge

\author{
Nathaniel J.~Smith\thanks{...} \\
School of Informatics\\
University of Edinburgh\\
Edinburgh, Scotland, UK\\
\texttt{nathaniel.smith@ed.ac.uk} \\
\AND
Noah D.~Goodman \\
Department of Psychology\\
Stanford University \\
Stanford, CA, USA \\
\texttt{ngoodman@stanford.edu} \\
\And
Michael C.~Frank \\
Department of Psychology \\
Stanford University \\
Stanford, CA, USA \\
\texttt{mcfrank@stanford.edu}}

@@

One option is to continue iterating such
strategies until reaching a fixed point equilibrium. While this
strategy guarantees that each agent will behave normatively given the
other agent's strategy, there is no guarantee that such strategies
will be near the system's global optimum. More importantly, there is a
great deal of
evidence that humans do not use such equilibrium strategies; their behavior in language
games (and in other games \cite{camerer2004}) can be well-modeled as
implementing $S_n$ or $L_n$ for some small $n$
\cite{frank2012}. Following this work, we recurse a finite (small)
number of times. The consequence is that one agent, implementing
$S_n$, is fully optimal with respect to the game, while the other,
implementing $L_{n-1}$, is only nearly optimal---off by a single
recursion.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data}$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
refers to successively stronger restrictions on the situations
compatible with the term. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating all of them, the listener is lead
to believe that ``some but not all'' is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.

>>>>

\author{
Nathaniel Smith\thanks{...} \\
School of Informatics\\
University of Edinburgh\\
Edinburgh, Scotland, UK\\
\texttt{nathaniel.smith@ed.ac.uk} \\
\AND
Noah D. Goodman \\
Department of Psychology\\
Stanford University \\
Stanford, CA, USA \\
\texttt{ngoodman@stanford.edu} \\
\And
Michael C. Frank \\
Department of Psychology \\
Stanford University \\
Stanford, CA, USA \\
\texttt{mcfrank@stanford.edu}}

@@

One option is to continue iterating such
strategies until reaching a fixed point equilibrium. While this
strategy guarantees that each agent will behave normatively given the
other agent's strategy, there is no guarantee that such strategies
will be near the system's global optimum. More importantly, is there much
evidence that humans do not use such equilibrium strategies; their behavior in language
games (and in other games \cite{camerer2004}) can be well-modeled as
implementing $S_n$ or $L_n$ for some small $n$
\cite{frank2012}. Following this work, we recurse a finite (small)
number of times. The consequence is that one agent, implementing
$S_n$, is fully optimal with respect to the game, while the other,
implementing $L_{n-1}$, is only nearly optimal---off by a single
recursion.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating all of them, the listener is lead
to believe that ``some but not all'' is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.
